---
title: "Home"
---

Welcome to Shirley's PM566 Final Project Website.

This project is a summary of analysis and figures related to a public health topic of Shirley's interest: mental health care and access.

Please go to this link to view the full report: \_\_\_.pdf

# Trends in Lack of Mental Health Care Since the COVID-19 Pandemic 

## Brief Description

Mental health is becoming an increasing concern since the 2019 coronavirus pandemic, with heightened stress, fear, and loneliness, along with elevated rates of "anxiety, anger, confusion, and posttraumatic symptoms" (Khan et al., 2022; Pakpour and Griffiths, 2020; Christoffersen et al., 2023). Mental health stigma and barriers to care exacerbate this issue. By identifying cohorts that may not be receiving care needed, barriers to access and resources may be more clearly outlined and show insight into the development of potentially more effective targeted therapies and care options to promote improved health of all populations.

This project intends to examine a few questions:

Is there a trend in individuals in the United States that need counseling or therapy for mental health disorders but did not receive relevant care? Is there a difference between different demographics (age, education status) and/or between different years (2022, 2021, 2020) and months within the year?

### Data

This data is acquired from the CDC (Centers for Disease Control and Prevention) Household Pulse Surveys, an experimental data system created by partnership between the National Center for Health Statistics (NCHC) and the U.S. Census Bureau along with other federal agencies. The intention was to "rapidly respond and provide relevant information about the impact of the coronavirus pandemic in the U.S. Data collection" beginning April 23, 2020. Specifically, this dataset is called "Mental Health Care in the Last 4 Weeks" where each four week period was a collection time point.

Please access more information through this link: https://www.cdc.gov/nchs/covid19/pulse/mental-health-care.htm

The dataset can also accessed through instructions described:

1.  *Have a `data` folder with either the dataset or instructions about how to acquire it. You should provide the instructions in a `README.md` file within that folder.*

## Visualizations  

```{r, echo=FALSE}
ggplot(age_subgrp_nocare_mean, aes(x = Subgroup, y = mean_value, fill = Subgroup)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Figure 1. Average Percent of Individuals Lacking Mental Health Counseling or Therapy by Age", 
    x = "Age Group", 
    y = "Average Percent"
    ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 12))
```

Figure 1. describes the average percent of individuals lacking mental health counseling or therapy in the past 4 weeks by age subgroup. A higher average percent indicates a greater lack of access (i.e., less care); a lower average percent indicates a lower lack of access (i.e., more care).

The ages 18-29 years old had the highest rates of lack of mental health care at 19.85%, and those 70-79 years old with the lowest rates at 3.26%.

```{r, echo=FALSE}
#reorder Subgroups on graph
custom_order <- c("Less than a high school diploma", "High school diploma or GED", "Some college/Associate's degree", "Bachelor's degree or higher")

edu_subgrp_nocare_mean$Subgroup <- factor(edu_subgrp_nocare_mean$Subgroup, levels = custom_order)

ggplot(edu_subgrp_nocare_mean, aes(x = Subgroup, y = mean_value, fill = Subgroup)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Figure 3. Average Percent of Individuals Lacking Mental Health Counseling or Therapy by Education Level", 
    x = "Education Level", 
    y = "Average Percent"
    ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 12))
```

Figure 3. describes the average percent of individuals lacking mental health counseling or therapy in the past 4 weeks by education subgroup.

There is no distinct trend observed in education subgroups with regards to lack of mental health care, with fluctuations across education level. Those with High school diploma or GED were at the lowest average percent at 8.75% and Some college/Associate's at the highest at 13.47%.

```{r, echo=FALSE}
ggplot(yr_nocare_mean, aes(x = year, y = mean_value)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Figure 5. Average Percent of Individuals Lacking Mental Health Counseling or Therapy per Year from 2020 to 2022",
    x = "Year",
    y = "Average Percent"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 12)) + 
  scale_x_continuous(breaks = unique(yr_nocare_mean$year), labels = unique(yr_nocare_mean$year))
```

Figure 5. describes the average percent of individuals lacking mental health counseling or therapy in the past 4 weeks by year.

In 2020, an average of 10.54% lacked mental health care, with an increase to 11.27% in 2021, and at 12.20% in 2022.

```{r, echo=FALSE}
# month_names <- data.frame(month = 1:12, month_name = month.name)
mo_nocare_mean$month <- factor(mo_nocare_mean$month, levels = 1:12, labels = month.name)

ggplot(mo_nocare_mean, aes(x = month, y = mean_value, group = 1)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Figure 6. Average Percent of Individuals Lacking Mental Health Counseling or Therapy per Month from 2020 to 2022",
    x = "Months",
    y = "Mean Value"
  ) +
  theme_minimal() + 
  # scale_x_discrete(
  #   breaks = month_names$month,
  #   labels = month_names$month_name)
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 12))
```

Figure 6. describes the average percent of individuals lacking mental health counseling or therapy in the past 4 weeks by month.

The overall month average percent trend appears to be positive from January to December, though noticeably large fluctuations from month to month are observed. May was observed to be the month with the lowest percent lack of mental health care at 9.86%, with December the highest at 12.27%.

## Conclusions

This data showed that since 2020, lack of mental health care has been on a generally increasing trend across all examined groups (Age, Education level), with mild differences. This is concerning but perhaps not unexpected given potential lag time between the start of the pandemic in 2019, leading to possible increase in incidence of mental health issues, time between symptoms and clinical diagnosis, stigma of mental health illness, and then time to seek and receive care. Certainly, insurance and other systemic health delivery aspects play a role as well. Multiple factors must be considered for the reason behind this increase, and future surveying and research are necessary.
